Gradual Soundness: Lessons from Static Python

### 2021-12-15 pj-7.1/paper.tex
pj: sec 2.0

  \paragraph{Importance}
  % Why does this work matter?
  \SP{} is the first language that improves performance across the board through
  the addition of sound types to an untyped language.
  %% Prior work on Nom and Dart 2 showed that concrete types are promising.
  Other language designers may wish to replicate its approach,
  especially those who maintain unsound type checkers for untyped languages,
  e.g., JavaScript.

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the
\SP{} language design in another context and measure its performance.
Second, we want to lend focus to researchers.
Some of the restrictions that \SP{} adopts may be useful for
other languages.
Other restrictions might be lifted by future research.

@@

@@

@@

@@

More notes and surprises:

\begin{itemize}
  \item
    Optional(T) is the only union. Others turn into Dynamic.
  \item
    TBD
\end{itemize}

@@

\begin{itemize}
  \item \textbf{First-order Typed Classes}.
    In order to benefit from types, class definitions must be defined at the top level of a module.
    Nested classes are supported by the language, but have type \code{Dynamic} according
    to the type checker.

  \item \textbf{No Overloading}.
    Methods cannot be overloaded. Each combination of a class
    and a name can refer to at most one method.
    This ensures that method resolution depends only on the
    type of the receiver object.
    % resolution does NOT depend on result type (T doesn't matter in `x:T = o.m()`)
    % Q. does resolution depend on args?

@@

  \item \textbf{Dynamic Checked Types}.
    A \code{CheckedDict} cannot be instantiated with \code{Dynamic}
    keys or values.
    [FILL] no desire, and makes runtime checks much simpler.
    %% NB can't do ChkDict[ChkDict[_,_],_] either but that's because ChkDict is unhashable for now

  \item \textbf{Recursive Types}.
    Sadly this is normal for Python type checkers.

>>>>

  \paragraph{Importance}
  % Why does this work matter?
  \SP{} is the first language that improves performance across the board through
  the addition of sound types to an untyped language.
  %% Prior work on Nom and Dart 2 showed that concrete types are promising.
  Other language designers may wish to replicate its approach,
  especially those who maintain unsound type checkers for untyped languages
  such as JavaScript.

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the
\SP{} language design.
In particular, the maintainers of optionally-typed languages
may find that adding transient and concrete types is a low-overhead
way to turn types into reliable claims.
Second, we want to lend focus to researchers.
Some of the restrictions that \SP{} adopts may be useful to
ground theoretical work.
Other restrictions might be lifted by future research.

@@

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
Relative to these sources, the \SP{} takes some interesting turns
because of the engineering context at Instagram.
The following contextual points are important to bear in mind as we present
notable aspects of the type system:
\begin{itemize}
  \item
    Performance is the bottom line.
    Unless \SP{} helps production code run faster, the team will be disbanded
    and reallocated.
    %% 2021-12-14: too extreme? and what about soundness?
  \item
    Instagram is / had been using Cython for performance.
    Cython compiles Python source code to C extension modules for
    a faster Python [CITE].
    Instagram had converted several performance-critical modules to Cython prior
    to starting work on \SP{}.
    \SP{} replaced them.
    (How does this help understanding type system highlights?)
  \item
    The Instagram codebase is already typed.
    Developers use Pyre [CITE], an optional type checker for Python,
    to write type annotations and to drive IDE tools (e.g., autocompletion).
  \item
    Q. why is the python so nominal? style guide? java background?
\end{itemize}

@@

%% 2021-12-14:
%% - anything to say about Any being a "meta" type in Python typing module?

\SP{} includes a dynamic type to accommodate untyped Python code.
Any un-annotated value or expression gets the dynamic type by default.
\SP{} does not make any predicitions / optimizations about dynamic code;
it assumes nothing and lets the Python runtime handle.

- lack of ann
- or, incorrect ann
- at toplevel, everything compatible
- diffirent rules inside data structures

@@

    Optional(T) is the only union. Others turn into Dynamic.

@@

\begin{itemize}
  \item \textbf{Recursive Types}.
    Sadly this is normal for Python type checkers.
\end{itemize}

\subsubsection{First-Order Functions and Classes}

Incidentally, \SP{} is a first-order type system.

... no user-facing type for functions
 imports / exports follow transient

... ditto for classes (see below)

  \item \textbf{First-order Typed Classes}.
    In order to benefit from types, class definitions must be defined at the top level of a module.
    Nested classes are supported by the language, but have type \code{Dynamic} according
    to the type checker.

@@

\begin{itemize}
  \item \textbf{No Overloading}.
    Methods cannot be overloaded. Each combination of a class
    and a name can refer to at most one method.
    This ensures that method resolution depends only on the
    type of the receiver object.
    % resolution does NOT depend on result type (T doesn't matter in `x:T = o.m()`)
    % Q. does resolution depend on args?

@@

%  \item \textbf{Dynamic Checked Types}.
%    A \code{CheckedDict} cannot be instantiated with \code{Dynamic}
%    keys or values.
%    [FILL] no desire, and makes runtime checks much simpler.
%    %% NB can't do ChkDict[ChkDict[_,_],_] either but that's because ChkDict is unhashable for now
